WASP-4 is Accelerating Toward the Earth

### 2020-02-09 paper/ms.tex
paper improvements

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TODO: make table 4 look beter

@@

%% Reintroduced the \received and \accepted commands from AASTeX v5.2.
%% Add "Submitted to " argument.
\received{\today}
\revised{---}
\accepted{---}
\submitjournal{AAS journals.}
\shorttitle{WASP-4 is accelerating towards the Sun}

@@

\title{WASP-4 is Accelerating Towards the Sun}
% ", Explaining its Decreasing Orbital Period"
% ": A Cautionary Tale in the Search for Tidal Orbital Decay"
% "An Explanation for WASP-4's Decreasing Orbital Period".
% "WASP-4's Decreasing Orbital Period: Solved."

@@

%
% contributing authors: alphabetical
%
\author[0000-0001-8638-0320]{A. W. Howard}
\affiliation{Cahill Center for Astrophysics, California Institute of
Technology, Pasadena, CA 91125, USA}
%
\author{S. B. Howell}
\affiliation{NASA Ames Research Center, Moffett Field, CA 94035, USA}
%
\author[0000-0002-0531-1073]{H. Isaacson}
\affiliation{Astronomy Department, University of California, Berkeley,
CA 94720, USA}
%
\author{H. Knutson}
\affiliation{Division of Geological and Planetary Sciences, California
Institute of Technology, Pasadena, CA 91125, USA}
%
\author{R. A. Matson}
\affiliation{NASA Ames Research Center, Moffett Field, CA 94035, USA}
%

%TODO FIXME: some sentences
\begin{abstract}
  The orbital period of the hot Jupiter WASP-4b appears to be
  decreasing at a rate of $-8.64 \pm 1.26~{\rm msec}\,{\rm yr}^{-1}$,
  based on transit-timing measurements spanning 12 years.  Proposed
  explanations for the period change include tidal orbital decay,
  apsidal precession, and acceleration of the system along the line of
  sight.  To investigate further, we performed new radial velocity
  measurements and speckle imaging of WASP-4.  The radial-velocity
  data show that the system is accelerating towards the Sun at a rate
  of $-0.0422\pm 0.0028\,{\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}$.  The
  associated Doppler effect should cause the apparent period to shrink
  at a rate of $-5.94 \pm 0.39~{\rm msec}\,{\rm yr}^{-1}$, comparable
  to the observed rate.  Thus, the observed change in the transit
  period is mostly or entirely produced by the line-of-sight
  acceleration of the system.  This acceleration is probably caused by
  a wide-orbiting companion of mass 10-200$\,M_{\rm Jup}$ and orbital
  distance 10-100$\,$AU, based on the magnitude of the radial-velocity
  trend and the non-detection of any companion in the speckle image.
  We expect that 1 out of 6 hot Jupiters will show apparent period
  changes of the same order of magnitude as WASP-4, based on the
  hot-Jupiter companion statistics of  \citet{knutson_friends_2014}.
  Continued radial velocity monitoring of hot Jupiters is therefore
  essential to distinguish the effects of tidal orbital decay or
  apsidal precession from line-of-sight acceleration.
\end{abstract}

@@

An empirical resolution might be possible through long-term timing of
transits and occultations, seeking evidence for changes in the orbital
period.  For instance, long-term transit timing and radial velocity
measurements for WASP-12b have revealed a secular decrease in the
period at a rate of $\approx$30 milliseconds per year, which has been
interpreted as the effect of tidal orbital decay
\citep{maciejewski_departure_2016,patra_2017,yee_orbit_2020}.

@@

Most of these data are identical to the data presented by
\citetalias{bouma_wasp4b_2019}.  Twenty-two new times reported by
\citet{southworth_transit_2019} are included.  These transits were
observed from the 3.58m New Technology Telescope and the Danish 1.54m
telescope at La Silla, Chile, and the South African Astronomical
Observatory 1.0m telescope.  Additional timing measurements were also
reported recently by \citet{baluev_2019}, based on a homogeneous
analysis of archival ground-based observations.  We included twelve of
their ``high quality'' transit times from the TRAPPIST telescope (six
transits), El~Sauce (four transits), and \citet{petrucci_no_2013}.
For TRAPPIST and El~Sauce, we verified with the original observers
that the timestamps were amenable to the appropriate barycentric and
leap second corrections (M.~Gillon, P.~Evans, priv{.} comm{.}).  We
omitted the fourteen remaining \citeauthor{baluev_2019}
ETD\footnote{\url{http://var2.astro.cz/ETD}} times due to ambiguity in
whether leap-second corrections had been performed.  We did not
include in our analysis the four occultation times tabulated by
\citetalias{bouma_wasp4b_2019}, because of the large timing
uncertainties and negligible statistical power.

@@

Once we saw that the new HIRES observations implied a highly
significant trend in the radial velocity, we sought independent
evidence for a wide-orbiting companion by performing speckle imaging
with the Zorro instrument on the Gemini South 8m
telescope\citep[see][and the instrument
web-pages\footnote{\url{www.gemini.edu/sciops/instruments/alopeke-zorro/}}]{scott_nessi_2018}.
Zorro is a dual-channel speckle interferometer employing narrow-band
filters centered at 562$\,$nm and 832$\,$nm.

@@

The resulting period derivative for the quadratic model is 
\begin{equation}
\dot{P}
  = - (2.74 \pm 0.28)\times 10^{-10}
  = - 8.64 \pm 1.26~{\rm msec}\,{\rm yr}^{-1}.
  \label{eq:dP_dt_obs}
\end{equation}
This agrees to within 1-$\sigma$ of the value reported by
\citet{southworth_transit_2019} ($\dot{P} = - 9.2 \pm 1.1~{\rm
ms}\,{\rm yr}^{-1}$).  It is 2.3-$\sigma$ larger than the rate of
period decrease reported by \citepalias[][; $- 12.6 \pm 1.2~{\rm
ms}\,{\rm yr}^{-1}$]{bouma_wasp4b_2019}, presumably because of the new
data from \citeauthor{southworth_transit_2019} and
\citeauthor{baluev_2019}. The other best-fitting model parameters are
reported in Table~3.

@@

Based on the available data, the companion causing the acceleration is
probably either a brown-dwarf or low mass star with an orbital
distance of $10$-$100\,{\rm AU}$ (Figure~\ref{fig:mass_sma}).  Given
such a mass, this companion may have influenced the orbital evolution
of the giant planet around WASP-4, as well as any other planets.  The
fact that most hot Jupiters have similar massive outer companions
\citep{knutson_friends_2014,bryan_statistics_2016} is compatible with
some of the high-eccentricity formation theories for hot Jupiters (see
\citealt{dawson_johnson_2018}).  Further radial velocity monitoring
should eventually reveal the orbital parameters and minimum mass of
WASP-4's companion.

@@

For hot Jupiters that have been monitored over baselines exceeding 10
years, secular changes in their orbital periods are currently being
constrained to a precision of $\lesssim$$10~{\rm msec}\,{\rm yr}^{-1}$
({\it e.g.}, K.~Patra et al{.} 2020, submitted).  {\bf Other citations
too?} This is roughly commensurate with the level of signal many outer
companions are expected to induce (Figure~\ref{fig:pdot_pop}).

@@

From newly acquired radial velocity measurements, we found that WASP-4
is accelerating towards the Earth at $\dot{\gamma} =
-0.0422^{+0.0028}_{-0.0027}\ {\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}$.
The corresponding Doppler effect predicts a period decrease
$\dot{\gamma} P/c$ of $-5.94 \pm 0.39~{\rm msec}\,{\rm yr}^{-1}$.  The
majority of the period decrease observed in transits ($\dot{P}= - 8.64
\pm 1.26~{\rm msec}\,{\rm yr}^{-1}$) is therefore explained by the
acceleration of the host star~---~leaving no evidence for tidal
orbital decay or apsidal precession.  The companion causing the
acceleration is most likely a brown dwarf or low-mass star with
semi-major axis between 10-100$\,$AU.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TODO: kishore reference
% TODO: rachel comments
% TODO: steve comments

@@

%% Reintroduced the \received and \accepted commands from AASTeX v5.2.
%% Add "Submitted to " argument.
\received{\today}
\revised{---}
\accepted{---}
\submitjournal{AAS journals.}
\shorttitle{WASP-4 is accelerating toward the Earth}

@@

\title{WASP-4 is Accelerating Toward the Earth}
% ", Explaining its Decreasing Orbital Period"
% ": A Cautionary Tale in the Search for Tidal Orbital Decay"
% "An Explanation for WASP-4's Decreasing Orbital Period".
% "WASP-4's Decreasing Orbital Period: Solved."

@@

%
% contributing authors: alphabetical
%
\author[0000-0001-8638-0320]{A. W. Howard}
\affiliation{Cahill Center for Astrophysics, California Institute of
Technology, Pasadena, CA 91125, USA}
%
\author{S. B. Howell}
\affiliation{NASA Ames Research Center, Moffett Field, CA 94035, USA}
%
\author[0000-0002-0531-1073]{H. Isaacson}
\affiliation{Astronomy Department, University of California, Berkeley,
CA 94720, USA}
%
\author{H. Knutson}
\affiliation{Division of Geological and Planetary Sciences, California
Institute of Technology, Pasadena, CA 91125, USA}
%
\author[0000-0001-7233-7508]{R. A. Matson}
\affiliation{U.S. Naval Observatory, Washington, DC 20392, USA}
%

\begin{abstract}
  The orbital period of the hot Jupiter WASP-4b appears to be
  decreasing at a rate of $-8.64 \pm 1.26~{\rm msec}\,{\rm yr}^{-1}$,
  based on transit-timing measurements spanning 12 years.  Proposed
  explanations for the period change include tidal orbital decay,
  apsidal precession, and acceleration of the system along the line of
  sight.  To investigate further, we performed new radial velocity
  measurements and speckle imaging of WASP-4.  The radial-velocity
  data show that the system is accelerating towards the Sun at a rate
  of $-0.0422\pm 0.0028\,{\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}$.  The
  associated Doppler effect should cause the apparent period to shrink
  at a rate of $-5.94 \pm 0.39~{\rm msec}\,{\rm yr}^{-1}$, comparable
  to the observed rate.  Thus, the observed change in the transit
  period is mostly or entirely produced by the line-of-sight
  acceleration of the system.  This acceleration is probably caused by
  a wide-orbiting companion of mass 10-200$\,M_{\rm Jup}$ and orbital
  distance 10-100$\,$AU, based on the magnitude of the radial-velocity
  trend and the non-detection of any companion in the speckle images.
  We expect that the orbital periods of 1 out of 6 hot Jupiters will
  appear to shrink at a rate similar to WASP-4b, based on the
  hot-Jupiter companion statistics of \citet{knutson_friends_2014}.
  Continued radial velocity monitoring of hot Jupiters is therefore
  essential to distinguish the effects of tidal orbital decay or
  apsidal precession from line-of-sight acceleration.
\end{abstract}

@@

An empirical resolution might be possible through long-term timing of
transits and occultations, seeking evidence for changes in the orbital
period.  For instance, long-term transit timing and radial velocity
measurements for WASP-12b have revealed a secular decrease in the
period at a rate of $\approx$30 milliseconds per year, which has been
interpreted as the effect of tidal orbital decay
\citep{maciejewski_departure_2016,patra_2017,maciejewski_planet-star_2018,yee_orbit_2020}.

@@

Most of these data are identical to the data presented by
\citetalias{bouma_wasp4b_2019}.  Twenty-two new times reported by
\citet{southworth_transit_2019} are included.  These transits were
observed from the 3.58m New Technology Telescope and the Danish 1.54m
telescope at La Silla, Chile, and the South African Astronomical
Observatory 1.0m telescope.  Additional timing measurements were also
reported recently by \citet{baluev_2019}, based on a homogeneous
analysis of archival ground-based observations.  We included twelve of
their ``high quality'' transit times from the TRAPPIST telescope (six
transits), El~Sauce (four transits), and \citet{petrucci_no_2013}.
For TRAPPIST and El~Sauce, we verified with the original observers
that the timestamps were amenable to the appropriate barycentric and
leap second corrections (M.~Gillon, P.~Evans, priv{.} comm{.}).  We
omitted the fourteen remaining 
ETD\footnote{\url{http://var2.astro.cz/ETD}} times from \citeauthor{baluev_2019} due to ambiguity in
whether leap-second corrections had been performed.  We did not
include in our analysis the four occultation times tabulated by
\citetalias{bouma_wasp4b_2019}, because of the large timing
uncertainties and negligible statistical power.

@@

Once we saw that the new HIRES observations implied a highly
significant trend in the radial velocity, we sought independent
evidence for a wide-orbiting companion by performing speckle imaging
with the Zorro instrument on the Gemini South 8m
telescope~\citep[see][and the instrument
web-pages\footnote{\url{www.gemini.edu/sciops/instruments/alopeke-zorro/}}]{scott_nessi_2018}.
Zorro is a dual-channel speckle interferometer employing narrow-band
filters centered at 562$\,$nm and 832$\,$nm.

@@

The resulting period derivative for the quadratic model is 
\begin{equation}
\dot{P}
  = - (2.74 \pm 0.28)\times 10^{-10}
  = - 8.64 \pm 1.26~{\rm msec}\,{\rm yr}^{-1}.
  \label{eq:dP_dt_obs}
\end{equation}
This agrees to within 1-$\sigma$ of the value reported by
\citet{southworth_transit_2019} ($\dot{P} = - 9.2 \pm 1.1~{\rm
ms}\,{\rm yr}^{-1}$).  It is 2.3-$\sigma$ larger than the rate of
period decrease reported by \citetalias{bouma_wasp4b_2019}
($- 12.6 \pm 1.2~{\rm ms}\,{\rm yr}^{-1}$), presumably because of the new
data from \citeauthor{southworth_transit_2019} and
\citeauthor{baluev_2019}. The other best-fitting model parameters are
reported in Table~3.

@@

Based on the available data, the companion causing the acceleration is
probably either a brown-dwarf or low mass star with an orbital
distance of $10$-$100\,{\rm AU}$ (Figure~\ref{fig:mass_sma}).  Given
such a mass, this companion may have influenced the orbital evolution
of the hot Jupiter orbiting WASP-4, as well as any other planets.  The
fact that most hot Jupiters have similar massive outer companions
\citep{knutson_friends_2014,bryan_statistics_2016} is compatible with
some of the high-eccentricity formation theories for hot Jupiters (see
\citealt{dawson_johnson_2018}).  Further radial velocity monitoring
should eventually reveal the orbital parameters and minimum mass of
WASP-4's companion.

@@

For hot Jupiters that have been monitored over baselines exceeding 10
years, secular changes in their orbital periods are currently being
constrained to a precision of $\lesssim$$10~{\rm msec}\,{\rm yr}^{-1}$
\citep{wilkins_searching_2017,maciejewski_planet-star_2018,baluev_2019,petrucci_discarding_2020,patra_2020}.
This is roughly commensurate with the level of signal many outer
companions are expected to induce (Figure~\ref{fig:pdot_pop}).

@@

From newly acquired radial velocity measurements, we found that WASP-4
is accelerating towards the Earth at $\dot{\gamma} =
-0.0422^{+0.0028}_{-0.0027}\ {\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}$.
The corresponding Doppler effect predicts a period decrease
$\dot{\gamma} P/c$ of $-5.94 \pm 0.39~{\rm msec}\,{\rm yr}^{-1}$.  The
majority of the period decrease observed in transits ($\dot{P}= - 8.64
\pm 1.26~{\rm msec}\,{\rm yr}^{-1}$) is therefore explained by the
acceleration of the host star~---~leaving no evidence for tidal
orbital decay or apsidal precession.  
A probabilistic analysis of the speckle imaging limits and the
radial velocity trend showed that the companion causing the
acceleration is most likely a brown dwarf or low-mass star with
semi-major axis between 10-100$\,$AU.
